\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-03-21 paper.tex
Structural updates addressing comments

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress, however, has led to a
% Controversial, perhaps, but true?
substantial gap opening between theory and practice.
Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
based on exhaustively detailing the ``events'' that occur
in the history of a sample,
are insufficient to describe the outputs of current methods.
Moreover, we argue that the underlying assumption
that all events can be known and precisely estimated
is fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).
Other subtly different concepts that we must be careful to distinguish
are the ``true'' ARG, % I don't see a better word for "true" here
describing the actual history of a sample of genomes,
and a ``population'' ARG which is
the true ARG of every individual in a population.
True ARGs are the underlying real history of a sample,
perfectly resolved into binary splits and mergers by
the cellular processes of meiosis and mitosis, regardless
of sampling density or population processes
(Appendix~\ref{sec-cell-lineages-and-args}).
Although population-scale true ARGs unquestionably exist,
they can also never be entirely known, in part because of a fundamental
lack of mutational information.
Even if mutation rate were high enough to uniquely identify every
recent branching point,
such a high rate would saturate the genome with mutations
and obscure deeper history.

% Population ARGs are here, now.
Population ARGs may seem fanciful, but the scale of
modern datasets makes it necessary for us to grapple with the idea.
The UK Biobank (UKB), for example, has
genotype data for around 500,000 humans~\citep{bycroft2018genome},
along with exome~\citep{backman2021exome} and
whole genome sequence~\citep{halldorsson2022sequences} data
for large subsets of the cohort.
UKB is just one of many such population-scale sequencing
projects~\citep[e.g.][]{turnbull2018hundred,
karczewski2020mutational,tanjo2021practical}.
Agricultural datasets are on a similar scale,
and also include dense multi-generational sampling
and near-perfect pedigree
information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
Recent advances have made it possible to actually
\emph{estimate} ARGs at population scale:
ARGs have already been inferred for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
While this new population-scale reality presents many
exciting opportunities, it also poses substantial challenges
to existing methodologies.

% What's the problem?
A major problem currently facing the field is that
classical mathematical formalisms and terminology
cannot adequately describe these vast inferred ARGs.
Fundamentally, these formalisms assume that an
ARG is known in complete detail
and are not suited to describing partial or approximate knowledge.
As we are actively inferring ARGs at the population scale,
and such ARGs can never be known in complete detail,
there is currently a substantial gap between our theoretical frameworks
and practical application.
The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
(Note that it is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.)
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

% This is nice to have, but probably won't work in the final version
% without section numbers. Let's deal with that when we get there I guess.
We begin in Section~\ref{sec-gARG} by providing a precise formal definition
of a gARG, illustrated by an example ARG embedded in pedigree.
We then provide a similar definition of the classical eARG
approach in Section~\ref{sec-eARG}, and consider its limitations
in the context of current datasets and research questions.
Following this, we discuss the important concept of ancestral material
in Section~\ref{sec-ancestral-material}, and how it relates
to the process of converting an eARG to a gARG.
We continue
in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
between an ARG and its local trees. Contrary to the
prevailing view, we show that a suitably encoded sequence of local trees contains
precisely as much information as the corresponding ARG.
The gARG encoding opens a rich new set of details about ARGs,
including the ideas of locally unary nodes
(Section~\ref{sec-locally-unary-edges}),
the levels of detail that can be represented in an ARG
(Section~\ref{sec-ARG-simplification}),
and the degrees of precision about recombination
that can be stored and we may seek to infer (Section~\ref{sec-precision}).
These ideas have important practical considerations,
which we illustrate by examining the qualitative properties of
ARGs inferred by four recent methods for a classical benchmark
dataset in Section~\ref{sec-example-inferred-args}.
We then discuss how the gARG framework can be
efficiently implemented in Section~\ref{sec-efficiency},
and finish with an assessment of the key challenges facing
the field in the Discussion.
Finally, the literature on ARGs is large and
confusing, and we attempt to clarify some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
a description of the Big and Little ARG stochastic processes
 (Appendix~\ref{sec-big-and-little-arg}),
a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
and a discussion of ARGs at an individual vs cell
lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

@@

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such exact estimates provide
for downstream applications.
Many applications work by examining local trees independently,
making detailed information about recombination events superfluous.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account.
Thus, while SIA takes advantage of information about local tree correlation,
it is in quite an indirect way, and
clearly much of the detail about recombination events in an ARG is lost.
% Mentioning tsdate here doesn't advance the
% narrative I think.
% The \texttt{tsdate} algorithm and related naive estimator
% of ancestral location explicitly use the gARG encoding
% to reason about ancestral nodes~\citep{wohns2022unified}.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the
coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
guo2022recombination},
which currently requires the details of all recombination
events to be known.

The advantages of a model-agnostic representation that naturally
incorporates uncertainty about the ordering of events in an ARG
are well-illustrated by~\cite{zhan2023towards}, who
inferred ARGs using millions of SARS-CoV-2 sequences from the
GISAID database~\citep{shu2017gisaid}.
In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
distinguish them. The reconstructed ARGs thus contain polytomies
and non-leaf sample nodes (sequences with descendants also present
in the dataset) many of which only have a single child (i.e.\ are ARG-unary).
Recombination is an important factor in the evolution of
SARS-CoV-2 \citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
% Tried to jazz this up a bit, was quite dry
and the inferred ARGs contain an unprecedented level of detail about
the combined processes of viral mutation and recombination.
Because parental sequences
are generally never sampled themselves, and often
a recombinant strain is the product of multiple recombination events,
uncertainty around this is captured by recording the ancestry
of each part of the recombinant sequence without arbitrarily assigning
times or orderings for these events.

@@

>>>>

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress has led to a diversity of ARG definitions and representations.

We show that a synthesis of these views can be achieved by thinking
in terms of genomes and transmission
intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
Moreover, we argue that the underlying assumption
that all events can be known and precisely estimated
is fundamentally unsuited to the realities
of modern, population-scale datasets.
We propose an alternative mathematical formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history}).

%Other subtly different concepts that we must be careful to distinguish
%are the ``true'' ARG, % I don't see a better word for "true" here
%describing the actual history of a sample of genomes,
%and a ``population'' ARG which is
%the true ARG of every individual in a population.
%True ARGs are the underlying real history of a sample,
%perfectly resolved into binary splits and mergers by
%%the cellular processes of meiosis and mitosis, regardless
%of sampling density or population processes
%(Appendix~\ref{sec-cell-lineages-and-args}).
%Although population-scale true ARGs unquestionably exist,
%they can also never be entirely known, in part because of a fundamental
%lack of mutational information.
%Even if mutation rate were high enough to uniquely identify every
%recent branching point,
%such a high rate would saturate the genome with mutations
%and obscure deeper history.

% Population ARGs are here, now.
%Population ARGs may seem fanciful, but the scale of
%modern datasets makes it necessary for us to grapple with the idea.
%The UK Biobank (UKB), for example, has
%genotype data for around 500,000 humans~\citep{bycroft2018genome},
%along with exome~\citep{backman2021exome} and
%whole genome sequence~\citep{halldorsson2022sequences} data
%for large subsets of the cohort.
%UKB is just one of many such population-scale sequencing
%projects~\citep[e.g.][]{turnbull2018hundred,
%karczewski2020mutational,tanjo2021practical}.
%Agricultural datasets are on a similar scale,
%and also include dense multi-generational sampling
%and near-perfect pedigree
%information~\citep[e.g.][]{hayes20191000,Ros-Freixedes2020}.
%Recent advances have made it possible to actually
%\emph{estimate} ARGs at population scale:
%ARGs have already been inferred for the 500,000
%humans in UKB~\citep{kelleher2019inferring,zhang2023biobank}
%and over a million SARS-CoV-2 genomes~\citep{zhan2023towards}.
%While this new population-scale reality presents many
%exciting opportunities, it also poses substantial challenges
%to existing methodologies.

% What's the problem?
%A major problem currently facing the field is that
%classical mathematical formalisms and terminology
%cannot adequately describe these vast inferred ARGs.
%Fundamentally, these formalisms assume that an
%ARG is known in complete detail
%and are not suited to describing partial or approximate knowledge.
%As we are actively inferring ARGs at the population scale,
%and such ARGs can never be known in complete detail,
%there is currently a substantial gap between our theoretical frameworks
%and practical application.

The breakthroughs in scale achieved by recent
methods~\citep[e.g.][]{kelleher2019inferring,speidel2019method,zhang2023biobank}
are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history.
% This is a digression from the maing point of the paragraph, so
% making it a parenthetical aside.
(Note that it is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.)
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

@@

% This is nice to have, but probably won't work in the final version
% without section numbers. Let's deal with that when we get there I guess.
% We begin in Section~\ref{sec-gARG} by providing a precise formal definition
% of a gARG, illustrated by an example ARG embedded in pedigree.
% We then provide a similar definition of the classical eARG
% approach in Section~\ref{sec-eARG}, and consider its limitations
% in the context of current datasets and research questions.
% Following this, we discuss the important concept of ancestral material
% in Section~\ref{sec-ancestral-material}, and how it relates
% to the process of converting an eARG to a gARG.
% We continue
% in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
% between an ARG and its local trees. Contrary to the
% prevailing view, we show that a suitably encoded sequence of local trees contains
% precisely as much information as the corresponding ARG.
% The gARG encoding opens a rich new set of details about ARGs,
% including the ideas of locally unary nodes
% (Section~\ref{sec-locally-unary-edges}),
% the levels of detail that can be represented in an ARG
% (Section~\ref{sec-ARG-simplification}),
% and the degrees of precision about recombination
% that can be stored and we may seek to infer (Section~\ref{sec-precision}).
% These ideas have important practical considerations,
% which we illustrate by examining the qualitative properties of
% ARGs inferred by four recent methods for a classical benchmark
% dataset in Section~\ref{sec-example-inferred-args}.
% We then discuss how the gARG framework can be
% efficiently implemented in Section~\ref{sec-efficiency},
% and finish with an assessment of the key challenges facing
% the field in the Discussion.
% Finally, the literature on ARGs is large and
% confusing, and we attempt to clarify some important aspects
% in appendices, including
% a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
% a description of the Big and Little ARG stochastic processes
%  (Appendix~\ref{sec-big-and-little-arg}),
% a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
% and a discussion of ARGs at an individual vs cell
% lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

\section{A diversity of structures}

3 PARAS. - enough outline to let readewr know what's in the appendices.

There are lots of new inference techniques: they all produce slightly different
structures, all of which can be though of a types of ARGs...

These give different levels of precision about recombination (see new Appendix X)

You can see this as different levels of detail on the same underlying truth. Here's a figure:

@@

%%% Pasted from previous para

@@

% % Do downstream applications actually make use of such full ARGs?
%Besides the inherent limitations that exist on inferring fully
%precise ARGs from data,
%we should also consider the value that such exact estimates provide
%for downstream applications.
%Many applications work by examining local trees independently,
%making detailed information about recombination events superfluous.
%For example, the \relate\ selection test~\citep{speidel2019method}
%obtains $p$-values by computing clade size probabilities conditional
%on the timing of coalescence events in a given local tree.
%In their method
%for estimating dispersal rates and the locations of genetic
%ancestors,
%\cite{osmond2021estimating} downsample trees along the genome
%so that they can be regarded as approximately independent.
%Similarly, \cite{fan2023likelihood} compute the likelihood
%of an ARG under a particular demographic model as the product
%over a sample of widely-separated local trees, assumed to be independent.
%The SIA method for detecting selection~\citep{hejase2022deep}
%encodes local trees as a set of lineage counts at discrete
%time intervals, and uses these as feature for a
%type of machine learning algorithm
%that takes ``temporal'' correlations into account.
%Thus, while SIA takes advantage of information about local tree correlation,
%it is in quite an indirect way, and
%clearly much of the detail about recombination events in an ARG is lost.

% Mentioning tsdate here doesn't advance the
% narrative I think.
% The \texttt{tsdate} algorithm and related naive estimator
% of ancestral location explicitly use the gARG encoding
% to reason about ancestral nodes~\citep{wohns2022unified}.
% The main application for fully precise ARGs thus far has been
% to compute a likelihood under the
% coalescent~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
% guo2022recombination},
% which currently requires the details of all recombination
% events to be known.

% The advantages of a model-agnostic representation that naturally
% incorporates uncertainty about the ordering of events in an ARG
% are well-illustrated by~\cite{zhan2023towards}, who
% inferred ARGs using millions of SARS-CoV-2 sequences from the
% GISAID database~\citep{shu2017gisaid}.
% In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
% sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
% distinguish them. The reconstructed ARGs thus contain polytomies
% and non-leaf sample nodes (sequences with descendants also present
% in the dataset) many of which only have a single child (i.e.\ are ARG-unary).
% Recombination is an important factor in the evolution of
% SARS-CoV-2 \citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
% % Tried to jazz this up a bit, was quite dry
% and the inferred ARGs contain an unprecedented level of detail about
% the combined processes of viral mutation and recombination.
% Because parental sequences
% are generally never sampled themselves, and often
% a recombinant strain is the product of multiple recombination events,
% uncertainty around this is captured by recording the ancestry
% of each part of the recombinant sequence without arbitrarily assigning
% times or orderings for these events.

@@

THE FIGURE OF INFERRED ARGS WAS HERE
